---
title: "cognitive_offloading_analysis_cogsci"
author: "sunny yu"
date: "2026-01-014"
output: html_document
---
# load dependencies 
```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(purrr)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(emmeans)
library(psych)
```

# read datasets: prediction_sample is the data collected from the prediction sample; completion_sample1 is data collected for the completion sample

```{r}
prediction_sample <- read_csv("/data/prediction_sample.csv")
completion_sample1 <- read_csv("/data/completion_sample1.csv")
```



### PRE-PROCESSING: clean data (prediction_sample and completion_sample for analysis)
## Pre-processing prediction sample: prediction_df
# We will include the following columns: ParticipantID, group (another participant or AI), taskID (6-character-long string, CATEGORY_QUESTION_GRADIENT), diffulcty (G1: easy, G2: difficult), prediction_target (independent vs. assisted), predicted_time
```{r}


prediction_sample <- prediction_sample %>%
  mutate(
    group = if_else(Condition == '1', "AI", "another_participant")
  )

task_map <- tibble::tibble(
  task_index = 1:4,
  task_col   = c("C1", "C2", "C3", "C4")
)

long_df <- map_dfr(seq_len(nrow(task_map)), function(i) {
  task_col <- task_map$task_col[[i]]
  k <- task_map$task_index[[i]]

  # Column names for this task index (k)
  ind_other_col <- paste0("task", k, "_independent_1")
  help_other_col <- paste0("task", k, "_help_1")

  ind_ai_col <- paste0("task", k, "_independent_AI_1")
  help_ai_col <- paste0("task", k, "_help_AI_1")

  prediction_sample %>%
    transmute(
      participantID = id,
      group = group,

      # 24-level taskID: e.g., "C1Q1G1"
      taskID = paste0(task_col, .data[[task_col]]),
      difficulty = str_sub(as.character(.data[[task_col]]), -2, -1),
      frequency = frequency,
      AIAS_1 = AIAS_1,
      AIAS_2 = AIAS_2,
      AIAS_3 = AIAS_3,
      AIAS_4 = AIAS_4,

      # choose the right numeric columns based on group
      predicted_independent = if_else(
        group == "AI",
        parse_number(as.character(.data[[ind_ai_col]])),
        parse_number(as.character(.data[[ind_other_col]]))
      ),
      predicted_assisted = if_else(
        group == "AI",
        parse_number(as.character(.data[[help_ai_col]])),
        parse_number(as.character(.data[[help_other_col]]))
      )
    ) %>%
    pivot_longer(
      cols = c(predicted_independent, predicted_assisted),
      names_to = "prediction_target",
      values_to = "predicted_time"
    ) %>%
    mutate(
      prediction_target = recode(
        prediction_target,
        predicted_independent = "independent",
        predicted_assisted    = "assisted"
      )
    )
})

prediction_df <- long_df %>%
  filter(!is.na(predicted_time)) %>%
  mutate(
    participantID = factor(participantID),
    taskID = factor(taskID),
    group = factor(group, levels = c("another_participant", "AI")),
    prediction_target = factor(prediction_target,
                               levels = c("independent", "assisted")),
    
  )

```


## pre-processing completion sample
# completion_df: We will include the following columns: participantID, condition (independnet vs. AI), need_for_cognition (average of NFC), taskID, difficulty, time, copypaste (1: copied and pasted, 0: not copied and pasted), NASA_TLX1-5/average
```{r}
# process data to include the following columns: condition (independent vs. AI), difficulty (G1 vs. G2), pid, taskID, completion time

task_map <- tibble::tibble(
  task_index = 1:4,
  task_col   = c("C1", "C2", "C3", "C4"),
  ind_time_col = c("task1_timer_Page Submit", "task2_timer_Page Submit",
                   "task3_timer_Page Submit", "task4_timer_Page Submit"),
  ai_time_col  = c("task1_timer_Page Submit.1", "task2_timer_Page Submit.1",
                   "task3_timer_Page Submit.1", "task4_timer_Page Submit.1")
)

# helper: safe numeric conversion (handles numeric, character, "NaN", etc.)
to_num <- function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) {
    x <- str_trim(x)
    x <- dplyr::na_if(x, "")
    x <- dplyr::na_if(x, "NaN")
    x <- dplyr::na_if(x, "NA")
    x <- dplyr::na_if(x, "N/A")
    return(readr::parse_number(x, na = c("", "NA", "NaN", "N/A")))
  }
  suppressWarnings(as.numeric(x))
}

# reverse code 1–7 (for TLX item 3 in your code)
rev7 <- function(x) ifelse(is.na(x), NA_real_, 8 - x)

# reverse code 1–5 (for NFC items 3 and 4)
rev5 <- function(x) ifelse(is.na(x), NA_real_, 6 - x)

completion_sample_nfc <- completion_sample1 %>%
  mutate(
    nfc1 = to_num(need_for_cognition_1),
    nfc2 = to_num(need_for_cognition_2),
    nfc3 = rev5(to_num(need_for_cognition_3)),  # reverse-coded
    nfc4 = rev5(to_num(need_for_cognition_4)),  # reverse-coded
    nfc5 = to_num(need_for_cognition_5),
    nfc6 = to_num(need_for_cognition_6),
    need_for_cognition = rowMeans(cbind(nfc1, nfc2, nfc3, nfc4, nfc5, nfc6), na.rm = TRUE)
  ) %>%
  select(-nfc1, -nfc2, -nfc3, -nfc4, -nfc5, -nfc6)

processed_df <- purrr::map_dfr(seq_len(nrow(task_map)), function(i) {
  task_col <- task_map$task_col[[i]]
  ind_col  <- task_map$ind_time_col[[i]]
  ai_col   <- task_map$ai_time_col[[i]]
  task_idx <- task_map$task_index[[i]]  # 1..4

  # build TLX column names for THIS TASK (task-first, item-second)
  tlx_ind_cols <- paste0("NASA-TLX", task_idx, "_", 1:5)
  tlx_ai_cols  <- paste0(tlx_ind_cols, ".1")  # AI has ".1" suffix

  copypaste_col <- paste0("prompt", task_idx, "_copypaste")

  completion_sample_nfc %>%   # <--- use the NFC-enriched df
    transmute(
      participantID = id,

      condition = dplyr::case_when(
        as.character(Condition) == "0" ~ "independent",
        as.character(Condition) == "1" ~ "AI",
        TRUE ~ NA_character_
      ),

      # carry NFC into long df
      need_for_cognition = need_for_cognition,
      task1_correctness = task1_correctness,
      task2_correctness = task2_correctness,
      task3_correctness = task3_correctness,
      task4_correctness = task4_correctness,
      final_answer1_correctness = final_answer1_correctness,
      final_answer2_correctness = final_answer2_correctness,
      final_answer3_correctness = final_answer3_correctness,
      final_answer4_correctness = final_answer4_correctness,
      frequency = frequency,
      AIAS_1 = AIAS_1,
      AIAS_2 = AIAS_2,
      AIAS_3 = AIAS_3,
      AIAS_4 = AIAS_4,

      # e.g., "C1Q1G1"
      taskID = paste0(task_col, .data[[task_col]]),

      # last 2 chars of the Ck value, e.g. "G1" / "G2"
      difficulty = str_sub(as.character(.data[[task_col]]), -2, -1),

      # completion time depends on Condition
      time = if_else(
        as.character(Condition) == "1",
        to_num(.data[[ai_col]]),
        to_num(.data[[ind_col]])
      ),

      copypaste = to_num(.data[[copypaste_col]]),

      # TLX items 1..5 for THIS TASK, chosen by Condition
      `NASA_TLX1` = if_else(
        as.character(Condition) == "1",
        to_num(.data[[tlx_ai_cols[1]]]),
        to_num(.data[[tlx_ind_cols[1]]])
      ),
      `NASA_TLX2` = if_else(
        as.character(Condition) == "1",
        to_num(.data[[tlx_ai_cols[2]]]),
        to_num(.data[[tlx_ind_cols[2]]])
      ),
      `NASA_TLX3` = rev7(if_else(   # reverse code item 3
        as.character(Condition) == "1",
        to_num(.data[[tlx_ai_cols[3]]]),
        to_num(.data[[tlx_ind_cols[3]]])
      )),
      `NASA_TLX4` = if_else(
        as.character(Condition) == "1",
        to_num(.data[[tlx_ai_cols[4]]]),
        to_num(.data[[tlx_ind_cols[4]]])
      ),
      `NASA_TLX5` = if_else(
        as.character(Condition) == "1",
        to_num(.data[[tlx_ai_cols[5]]]),
        to_num(.data[[tlx_ind_cols[5]]])
      ),

      NASA_TLX_average = rowMeans(
        cbind(`NASA_TLX1`, `NASA_TLX2`, `NASA_TLX3`, `NASA_TLX4`, `NASA_TLX5`),
        na.rm = TRUE
      )
    )
})

completion_df <- processed_df %>%
  filter(!is.na(time), !is.na(taskID), !is.na(condition), !is.na(difficulty)) %>%
  mutate(
    participantID = factor(participantID),
    taskID = factor(taskID),

    # create task_category BEFORE filtering
    task_category = substr(as.character(taskID), 1, 2),

    condition = factor(condition, levels = c("independent", "AI")),
    difficulty = factor(difficulty, levels = c("G1", "G2"))
  ) %>%
  filter(
    dplyr::case_when(
      condition == "independent" & task_category == "C1" ~ task1_correctness == 1,
      condition == "independent" & task_category == "C2" ~ task2_correctness == 1,
      condition == "independent" & task_category == "C3" ~ task3_correctness == 1,
      condition == "independent" & task_category == "C4" ~ task4_correctness == 1,

      condition == "AI" & task_category == "C1" ~ final_answer1_correctness == 1,
      condition == "AI" & task_category == "C2" ~ final_answer2_correctness == 1,
      condition == "AI" & task_category == "C3" ~ final_answer3_correctness == 1,
      condition == "AI" & task_category == "C4" ~ final_answer4_correctness == 1,

      TRUE ~ FALSE
    )
  )

```


## concatendate prediction and completion into combined_df
# added column: type (prediction vs. completion)

```{r}
# remove another participant from prediction_df (since we're only interested in data for AI vs. independent)

prediction_df_filtered <- prediction_df %>%
  filter(group != "another_participant") %>%   # keep AI group only
  mutate(
    condition = recode(prediction_target,
                       assisted = "AI",
                       independent = "independent"),
    difficulty = str_sub(as.character(taskID), -2, -1)
  ) %>%
  select(-prediction_target)

# Prepare prediction_sample
prediction_sample_prepped <- prediction_df_filtered %>%
  rename(time = predicted_time) %>%
  mutate(type = "prediction")

# Prepare completion_sample
completion_sample_prepped <- completion_df %>%
  mutate(type = "completion")

# Concatenate
combined_df <- bind_rows(
  prediction_sample_prepped,
  completion_sample_prepped,
) %>%
 mutate(
    task_category = substr(taskID, 1, 2),
    task_category = factor(task_category)
 )
```


#____________________________________________

### Analysis 
## Hypothesis 1: predicted time for other > self 
# model
```{r}

m1 <- lmer(
  predicted_time ~ prediction_target * group +
    (1 + prediction_target | participantID) +
    (1 | taskID),
  data = prediction_df
)
summary(m1)



```
# independent vs. assisted for AI
```{r}
# set group to AI to examine simple effects at prediction_target
emm_ai <- emmeans(m1, ~ prediction_target, at = list(group = "AI"))
contrast(emm_ai, method = "pairwise")

```
# independent vs. assisted for another participant
```{r}
emm_another_participant <- emmeans(m1, ~ prediction_target, at = list(group = "another_participant"))
contrast(emm_another_participant, method = "pairwise")
```

```{r}
# contrast between the two
emm <- emmeans(m1, ~ prediction_target * group)

contrast(
  emm,
  interaction = "pairwise",
  by = NULL
)
```

# category variations

```{r}

prediction_df <- prediction_df %>%
  mutate(
    task_category = substr(taskID, 1, 2),
    task_category = factor(task_category)
  )


m_cat_int <- lmer(
  predicted_time ~ prediction_target * group +
    prediction_target * task_category +
    (1 + prediction_target | participantID) +
    (1 | taskID),
  data = prediction_df
)

```

```{r}
# category variation for the AI condition
emm_ai_cat <- emmeans(m_cat_int, ~ prediction_target | task_category, at = list(group = "AI"))
contrast(emm_ai_cat, method = "pairwise")

```

```{r}
# category variation for the "another participant" condition
emm_another_participant <- emmeans(m_cat_int, ~ prediction_target | task_category, at = list(group = "another_participant"))
contrast(emm_another_participant, method = "pairwise")
```


# difficulty variation
```{r}

m_difficulty_int <- lmer(
  predicted_time ~ prediction_target * group +
    prediction_target * difficulty +
    (1 + prediction_target | participantID) +
    (1 | taskID),
  data = prediction_df
)

```


# difficulty variation for AI
```{r}
emm_ai_diff <- emmeans(m_difficulty_int, ~ prediction_target | difficulty, at = list(group = "AI"))
contrast(emm_ai_diff, method = "pairwise")
```


#difficulty variation for another participant
```{r}
emm_another_participant_diff <- emmeans(m_difficulty_int, ~ prediction_target | difficulty, at = list(group = "another_participant"))
contrast(emm_another_participant_diff, method = "pairwise")

```


```{r}
# task-level differences

m_task <- lmer(
  predicted_time ~ prediction_target * group * taskID +
    (1 | participantID),
  data = prediction_df
)


emm_ind <- emmeans(m_task, ~ prediction_target | taskID,
                   at = list(group = "AI"))

contrast(emm_ind, "pairwise", adjust = "holm")


emm_ap <- emmeans(m_task, ~ prediction_target | taskID,
                  at = list(group = "another_participant"))

contrast(emm_ap, "pairwise", adjust = "holm")



```


## INDEPENDENT vs. AI
```{r}
# model: 

m <- lmer(
  time ~ condition  + difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = completion_df
)
summary(m)



```



```{r}
m <- lmer(
  time ~ condition*difficulty +
    (1 | participantID) +
   (1 | taskID),
  data = completion_df
)
summary(m)

emm <- emmeans(m, ~ condition | difficulty)

contrast(emm, "pairwise", adjust = "holm")
```
# predicting nasa-tlx with nfc as a variable
```{r}
m <- lmer(
  NASA_TLX_average ~ condition*need_for_cognition  + difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = completion_df
)
summary(m)
```




```{r}
m <- lmer(
  time ~ condition*taskID*difficulty +
    (1 | participantID),
  data = completion_df
)
summary(m)

emm <- emmeans(m, ~ condition | taskID)

contrast(emm, "pairwise", adjust = "holm")
```


## CALIBRATION ANALYSIS: predicted vs. actual for independet and for AI

```{r}

# model: 

m_calibration <- lmer(
  time ~ condition * type  + difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = combined_df
)
summary(m_calibration)


```



```{r}
# simple effect of type for the independent condition (actual vs. predicted )
emm_indep <- emmeans(
  m_calibration,
  ~ type ,
  at = list(condition = "independent")
)

contrast(emm_indep, method = "pairwise", adjust = "holm")
```





```{r}
# simple effect of type for the AI condition (actual vs. predicted )
emm_AI <- emmeans(
  m_calibration,
  ~ type ,
  at = list(condition = "AI")
)

contrast(emm_AI, method = "pairwise", adjust = "holm")
```


```{r}
# task-category variations

# model
m_cat <- lmer(
  time ~ condition * type  + 
    type * task_category + 
    difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = combined_df
)

emm_indep <- emmeans(
  m_cat,
  ~ type | task_category,
  at = list(condition = "independent")
)

contrast(emm_indep, method = "pairwise", adjust = "holm")


emm_AI <- emmeans(
  m_cat,
  ~ type | task_category,
  at = list(condition = "AI")
)

contrast(emm_AI, method = "pairwise", adjust = "holm")


```
# difficulty variation
```{r}
# difficulty variations
m_diff <- lmer(
  time ~ condition * type +
    type* difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = combined_df
)

emm_indep <- emmeans(
  m_diff,
  ~ type | difficulty,
  at = list(condition = "independent")
)

contrast(emm_indep, method = "pairwise", adjust = "holm")


emm_AI <- emmeans(
  m_diff,
  ~ type | difficulty,
  at = list(condition = "AI")
)

contrast(emm_AI, method = "pairwise", adjust = "holm")
```

```{r}
# task-level variations

m_task <- lmer(
  time ~ condition * type * taskID +
    (1 | participantID),
  data = combined_df
)


emm_ind <- emmeans(m_task, ~ type | taskID,
                   at = list(condition = "independent"))

contrast(emm_ind, "pairwise", adjust = "holm")


emm_ai <- emmeans(m_task, ~ type | taskID,
                  at = list(condition = "AI"))

contrast(emm_ai, "pairwise", adjust = "holm")


```

## interaction effect between offloading and difficulty levels 
```{r}
# filter back to just the completion data
completion_df <- combined_df %>%
  dplyr::filter(type == "completion")

m_interaction <- lmer(
  time ~ condition * difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = completion_df,
)
summary(m_interaction)

```

## NASA-TLX: for the models above, change dv from time to NASA-TLX

```{r}
# correlation between completion time and NASA-TLX
cols <- c("time", "NASA_TLX1", "NASA_TLX2", "NASA_TLX3", "NASA_TLX4", "NASA_TLX5", "NASA_TLX_average")
completion_df %>% select(all_of(cols))
completion_df <- completion_df %>%
  mutate(across(all_of(cols), as.numeric))

corr.test(completion_df %>% select(all_of(cols)),
    method = "pearson", adjust = "holm")
```
# NASA-TLX overall
```{r}
# the most basic global model

m_global <- lmer(
  NASA_TLX_average ~ condition  + difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = completion_df,
)
summary(m_global)

```

# task category variation for NASA-TLX
```{r}
m_interaction_tc <- lmer(
  NASA_TLX_average ~ condition * difficulty * task_category +
         (1 | participantID) +
         (1 | taskID),
  data = completion_df
)
emm_tc <- emmeans(m_interaction_tc, ~ condition | task_category)
contrast(emm_tc, method = "pairwise", adjust = "none")
```

# difficulty variation for NASA-TLX
```{r}
emm_diff <- emmeans(m_interaction_tc, ~ condition | difficulty)
contrast(emm_diff, method = "pairwise", adjust = "none")
```


# per-task variation for NASA-TLX
```{r}
m_task <- lmer(
  NASA_TLX_average ~ condition * taskID +
         (1 | participantID),
  data = completion_df
)

emm <- emmeans(m_task, ~ condition | taskID)
contrast(emm, method = "pairwise", adjust = "none")

```


### EXPLORATORY ANALYSIS
# pre-processing: stated_preference (1: prefer assistance (AI or another participant), 0: prefer independent completion)
```{r}

# ---- helpers for NFC ----
to_num <- function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) {
    x <- str_trim(x)
    x <- dplyr::na_if(x, "")
    x <- dplyr::na_if(x, "NaN")
    x <- dplyr::na_if(x, "NA")
    x <- dplyr::na_if(x, "N/A")
    return(readr::parse_number(x, na = c("", "NA", "NaN", "N/A")))
  }
  suppressWarnings(as.numeric(x))
}

# reverse code (for a scale from 1–5)
rev5 <- function(x) ifelse(is.na(x), NA_real_, 6 - x)

prediction_sample <- prediction_sample %>%
  mutate(
    group = if_else(Condition == "1", "AI", "another_participant"),

    # ---- NEW: need for cognition (row-wise) ----
    nfc1 = to_num(need_for_cognition_1),
    nfc2 = to_num(need_for_cognition_2),
    nfc3 = rev5(to_num(need_for_cognition_3)),  # reverse-coded
    nfc4 = rev5(to_num(need_for_cognition_4)),  # reverse-coded
    nfc5 = to_num(need_for_cognition_5),
    nfc6 = to_num(need_for_cognition_6),
    need_for_cognition = rowMeans(cbind(nfc1, nfc2, nfc3, nfc4, nfc5, nfc6), na.rm = TRUE)
  ) %>%
  select(-nfc1, -nfc2, -nfc3, -nfc4, -nfc5, -nfc6)

# Map tasks to their category column AND the correct "choice" columns
task_map <- tibble::tibble(
  task_index = 1:4,
  task_col   = c("C1", "C2", "C3", "C4"),
  choice_other_col = c("task1_choice", "task2_choice", "task3_choice", "task4_choice"),
  choice_ai_col    = c("task1_choice_AI", "task2_choice_AI", "task3_choice_AI", "task4_choice_AI")
)

long_df <- map_dfr(seq_len(nrow(task_map)), function(i) {
  task_col <- task_map$task_col[[i]]
  k <- task_map$task_index[[i]]

  choice_other_col <- task_map$choice_other_col[[i]]
  choice_ai_col    <- task_map$choice_ai_col[[i]]

  # Column names for predicted-time inputs
  ind_other_col  <- paste0("task", k, "_independent_1")
  help_other_col <- paste0("task", k, "_help_1")
  ind_ai_col     <- paste0("task", k, "_independent_AI_1")
  help_ai_col    <- paste0("task", k, "_help_AI_1")

  prediction_sample %>%
    transmute(
      participantID = id,
      group = group,

      # ---- NEW: carry NFC through ----
      need_for_cognition = need_for_cognition,
      nfc1 = need_for_cognition_1,
           nfc2 = need_for_cognition_2,
           nfc3 = need_for_cognition_3,
           nfc4 = need_for_cognition_4,
           nfc5 = need_for_cognition_5,
      frequency = frequency,
      AIAS_1 = AIAS_1,
        AIAS_2 = AIAS_2,
        AIAS_3 = AIAS_3,
        AIAS_4 = AIAS_4,
      

      # 24-level taskID: e.g., "C1Q1G1"
      taskID = paste0(task_col, .data[[task_col]]),

      # task difficulty from task code (your original behavior)
      difficulty_level = str_sub(as.character(.data[[task_col]]), -2, -1),

      # predicted times (choose the right numeric columns based on group)
      predicted_independent = if_else(
        group == "AI",
        parse_number(as.character(.data[[ind_ai_col]])),
        parse_number(as.character(.data[[ind_other_col]]))
      ),
      predicted_assisted = if_else(
        group == "AI",
        parse_number(as.character(.data[[help_ai_col]])),
        parse_number(as.character(.data[[help_other_col]]))
      ),

      # stated preference: pick the right choice column based on group
      stated_choice_raw = if_else(
        group == "AI",
        as.character(.data[[choice_ai_col]]),
        as.character(.data[[choice_other_col]])
      ),
      stated_preference = case_when(
        str_detect(stated_choice_raw, "^Independently") ~ 0L,
        str_detect(stated_choice_raw, "^With") ~ 1L,
        TRUE ~ NA_integer_
      )
    ) %>%
    pivot_longer(
      cols = c(predicted_independent, predicted_assisted),
      names_to = "prediction_target",
      values_to = "predicted_time"
    ) %>%
    mutate(
      prediction_target = recode(
        prediction_target,
        predicted_independent = "independent",
        predicted_assisted    = "assisted"
      )
    )
})

stated_preference_df <- long_df %>%
  filter(!is.na(predicted_time)) %>%
  mutate(
    participantID = factor(participantID),
    taskID = factor(taskID),
    group = factor(group, levels = c("another_participant", "AI")),
    prediction_target = factor(prediction_target, levels = c("independent", "assisted")),
    difficulty_level = factor(difficulty_level),
    stated_preference = as.integer(stated_preference),
    frequency = frequency,
      AIAS_1 = AIAS_1,
      AIAS_2 = AIAS_2,
    AIAS_3 = AIAS_3,
      AIAS_4 = AIAS_4,
  )

# normalize NFC
stated_preference_df <- stated_preference_df %>%
  mutate(
    need_for_cognition_c = scale(need_for_cognition, center = TRUE, scale = TRUE)
  )

```


## stated preference

# prediction target and predicted preference
```{r}
m_sp <- glmer(
  stated_preference ~ group*difficulty_level +
    (1 + prediction_target | participantID) +
    (1 | taskID),
  data = stated_preference_df,
  family = binomial(link = "logit")
)
summary(m_sp)


emm <- emmeans(
  m_sp,
  ~ difficulty_level,
  type = "response"  # converts log-odds → probabilities
)

emm

```



## copying and pasting behaviors

```{r}

m_interaction <- lmer(
  time ~ condition * copypaste + difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = completion_df,
)
summary(m_interaction)

# how copying and pasting affects time

emm <- emmeans(m_interaction, ~ copypaste,
                  at = list(condition = "AI"))

contrast(emm, "pairwise", adjust = "holm")
```
# how copypaste affects subjective effort
```{r}
m_cp_nasa <- lmer(
  NASA_TLX_average ~ condition * copypaste + difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = completion_df,
)

emm_cp_nasa <- emmeans(m_cp_nasa, ~ copypaste,
                  at = list(condition = "AI"))
contrast(emm_cp_nasa, "pairwise", adjust = "holm")
```

## decomposing completion time (code used for pre-processing in Python)
# df structure: time_type (processing_time vs. prompting_time); time; generation_time (this is minimal, can ignore)

```{r}
# load decomp_df (df with data about decomposition of each component)

# helper
to_num <- function(x) {
  # Works for numeric, character, factors; converts "NaN"/"NA"/"" to NA
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) {
    x <- dplyr::na_if(x, "")
    x <- dplyr::na_if(x, "NaN")
    x <- dplyr::na_if(x, "NA")
    x <- dplyr::na_if(x, "N/A")
  }
  suppressWarnings(as.numeric(x))
}

# ---- task map ----
task_map <- tibble(
  task_index = 1:4,
  task_category = c("C1", "C2", "C3", "C4"),
  prompting_col   = paste0("task", 1:4, "_prompting_time"),
  generation_col  = paste0("task", 1:4, "_model_generation_time"),
  processing_col  = paste0("task", 1:4, "_processing_time")
)

processed_df <- map_dfr(seq_len(nrow(task_map)), function(i) {

  tm <- task_map[i, ]

  completion_sample_time_decomp %>%
    filter(as.character(Condition) == '1') %>%
    transmute(
      participantID = id,
      condition = "AI",

      # task category (C1–C4)
      task_category = tm$task_category,

      # e.g., C1Q1G1
      taskID = paste0(
        tm$task_category,
        .data[[tm$task_category]]
      ),

      # extract G1 / G2 from Ck column
      difficulty = str_sub(
        as.character(.data[[tm$task_category]]),
        -2, -1
      ),

      prompting_time  = to_num(.data[[tm$prompting_col]]),
      generation_time = to_num(.data[[tm$generation_col]]),
      processing_time = to_num(.data[[tm$processing_col]])
    )
})

decomp_df <- processed_df %>%
  filter(
    !is.na(prompting_time),
    !is.na(generation_time),
    !is.na(processing_time),
    !is.na(taskID),
    !is.na(difficulty)
  ) %>%
  mutate(
    participantID = factor(participantID),
    taskID = factor(taskID),
    task_category = factor(task_category, levels = c("C1", "C2", "C3", "C4")),
    difficulty = factor(difficulty, levels = c("G1", "G2")),
    condition = factor(condition, levels = "AI")
  )


decomp_df <- decomp_df %>%
  pivot_longer(
    cols = c(prompting_time, processing_time),
    names_to = "time_type",
    values_to = "time"
  ) %>%
  mutate(
    time_type = factor(
      time_type,
      levels = c("prompting_time", "processing_time")
    )
  )




```


# global difference
```{r}
m_global <- lmer(
  time ~ time_type +
    (1 | participantID) +
    (1 | taskID),
  data = decomp_df
)
summary(m_global)
```


```{r}
m_diff <- lmer(
  time ~ time_type * difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = decomp_df
)

emm <- emmeans(m_diff, ~ time_type | difficulty)
contrast(emm, "pairwise")

```
# task category
```{r}

m_cat <- lmer(
  time ~ time_type * task_category +
    (1 | participantID) +
    (1 | taskID),
  data = decomp_df
)


emm_cat <- emmeans(m_cat, ~ time_type | task_category)
contrast(emm_cat, "pairwise", adjust = "holm")


```
# for specific tasks

```{r}
m_cat <- lmer(
  time ~ time_type * taskID +
    (1 | participantID),
  data = decomp_df
)


emm_cat <- emmeans(m_cat, ~ time_type | taskID)
contrast(emm_cat, "pairwise", adjust = "holm")
```



## individual differences (Need for Cognition)
```{r}
# how need_for_cognition predicts copypaste
completion_ai_df <- completion_df %>%
  dplyr::filter(condition == "AI")

m_interaction <- lmer(
  copypaste ~  need_for_cognition + difficulty +
    (1 | participantID) +
    (1 | taskID),
  data = completion_ai_df,
)

summary(m_interaction)

```



# individual traits analysis
```{r}
# NFC and predicted time: does participants lower in NFC (lazier) predict a greater effect of offloading 

diff_df <- stated_preference_df %>%
  select(participantID, taskID, group, difficulty_level,
         prediction_target, predicted_time, nfc4) %>%
  pivot_wider(
    names_from = prediction_target,
    values_from = predicted_time
  ) %>%
  mutate(
    delta_predicted_time = assisted - independent,
    nfc4 = scale(nfc4, TRUE, TRUE)
  )

m_nfc_diff <- lmer(
  delta_predicted_time ~ nfc4 * group +
    (1 | participantID) +
    (1 | taskID),
  data = diff_df
)
summary(m_nfc_diff)

```




```{r}


library(stringr)

diff_df <- stated_preference_df %>%
  select(participantID, taskID, group, difficulty_level,
         prediction_target, predicted_time, frequency,
         AIAS_1, AIAS_2, AIAS_3, AIAS_4) %>%
  pivot_wider(
    names_from = prediction_target,
    values_from = predicted_time
  ) %>%
  mutate(
    delta_predicted_time = assisted - independent,
    frequency = factor(frequency),
    group = factor(group),
    AIAS_1_num = as.numeric(str_extract(AIAS_1, "\\d+$")),
    AIAS_2_num = as.numeric(str_extract(AIAS_2, "\\d+$")),
    AIAS_3_num = as.numeric(str_extract(AIAS_3, "\\d+$")),
    AIAS_4_num = as.numeric(str_extract(AIAS_4, "\\d+$"))
  )

m_nfc_diff_num <- lmer(
  delta_predicted_time ~ AIAS_4_num * group +
    (1 | participantID) +
    (1 | taskID),
  data = diff_df
)

summary(m_nfc_diff_num)
```



### PLOTTING CODE
# Figure 2: predicted vs. actual for the 2 conditions 
```{r}

# Bar plot: condition (independent vs AI) × type (prediction vs completion)
# with 95% CI bars (± 1.96 * SE) and significance brackets:
# - within independent: prediction vs completion = n.s.
# - within AI: prediction vs completion = significant


library(dplyr)
library(ggplot2)

# --- clean factors / labels ---
plot_df <- combined_df %>%
  mutate(
    condition = factor(condition, levels = c("independent", "AI")),
    type = factor(type, levels = c("prediction", "completion"),
                  labels = c("Predicted time", "Actual time"))
  )

# --- summarize for bars + CI (± 1.96 * SE) ---
sum_df <- plot_df %>%
  group_by(condition, type) %>%
  summarize(
    n = sum(!is.na(time)),
    mean_time = mean(time, na.rm = TRUE),
    se = sd(time, na.rm = TRUE) / sqrt(n),
    ci = 1.96 * se,
    .groups = "drop"
  )

# --- significance tests within each condition (prediction vs completion) ---
# This is a simple default. If you have participantID/taskID, replace with a mixed model.
p_ind <- t.test(time ~ type, data = filter(plot_df, condition == "independent"))$p.value
p_ai  <- t.test(time ~ type, data = filter(plot_df, condition == "AI"))$p.value

label_from_p <- function(p) {
  if (is.na(p)) return("n/a")
  if (p < 0.001) return("***")
  if (p < 0.01)  return("**")
  if (p < 0.05)  return("*")
  return("n.s.")
}

sig_df <- sum_df %>%
  group_by(condition) %>%
  summarize(
    y = max(mean_time + ci, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    y = y + 0.05 * max(sum_df$mean_time + sum_df$ci, na.rm = TRUE),  # small padding
    label = c(label_from_p(p_ind), label_from_p(p_ai))
  )

# helper to get dodge positions for bracket endpoints
dodge_w <- 0.8
get_x <- function(cond, type_label) {
  # ggplot uses discrete x positions 1,2,... for condition,
  # and dodges by type within each x.
  # With two levels of type, the dodged x offsets are approx ± dodge_w/4
  base <- as.numeric(factor(cond, levels = levels(sum_df$condition)))
  offset <- ifelse(type_label == "Predicted time", -dodge_w/4, dodge_w/4)
  base + offset
}

sig_df <- sig_df %>%
  rowwise() %>%
  mutate(
    x1 = get_x(condition, "Predicted time"),
    x2 = get_x(condition, "Actual time"),
    y1 = y,
    y2 = y + 0.03 * max(sum_df$mean_time + sum_df$ci, na.rm = TRUE) # bracket height
  ) %>%
  ungroup()

# --- plot ---
pdf("figure2_cogsci.pdf", width = 8, height = 6)
ggplot(sum_df, aes(x = condition, y = mean_time, fill = type)) +
  geom_col(position = position_dodge(width = dodge_w), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_time - ci, ymax = mean_time + ci),
    position = position_dodge(width = dodge_w),
    width = 0.18,
    linewidth = 0.6
  ) +
  # significance brackets per condition (prediction vs completion)
  geom_segment(
    data = sig_df,
    aes(x = x1, xend = x2, y = y2, yend = y2),
    inherit.aes = FALSE,
    linewidth = 0.6
  ) +
  geom_segment(
    data = sig_df,
    aes(x = x1, xend = x1, y = y1, yend = y2),
    inherit.aes = FALSE,
    linewidth = 0.6
  ) +
  geom_segment(
    data = sig_df,
    aes(x = x2, xend = x2, y = y1, yend = y2),
    inherit.aes = FALSE,
    linewidth = 0.6
  ) +
  geom_text(
    data = sig_df,
    aes(x = (x1 + x2) / 2, y = y2, label = label),
    inherit.aes = FALSE,
    vjust = -0.4,
    size = 4
  ) +
  scale_fill_discrete(name = NULL) +  # legend shows "Predicted time" / "Completion time"
  labs(
    x = NULL,
    y = "Time",
  ) +
  theme_classic(base_size = 20) +
  theme(legend.position = "top")

dev.off()

```




## Figure 4a: forest plot for NASA-TLX difference per task, ordered by effect size (left)


```{r}


library(lme4)
library(emmeans)
library(dplyr)
library(ggplot2)
library(forcats)

completion_df <- completion_df %>%
  mutate(
    condition = factor(condition, levels = c("independent", "AI")),
    taskID    = factor(taskID)
  )

# --- model ---
m_task <- lmer(
  NASA_TLX_average ~ condition * taskID + (1 | participantID),
  data = completion_df
)

# --- emmeans contrast: independent - AI ---
emm <- emmeans(m_task, ~ condition | taskID)
ct  <- contrast(emm, method = list("independent - AI" = c(1, -1)), adjust = "none")

# --- summary with CIs (robust across emmeans versions) ---
ct_df <- as.data.frame(summary(ct, infer = c(TRUE, TRUE)))

get_ci_cols <- function(df) {
  if (all(c("lower.CL", "upper.CL") %in% names(df))) return(c("lower.CL", "upper.CL"))
  if (all(c("asymp.LCL", "asymp.UCL") %in% names(df))) return(c("asymp.LCL", "asymp.UCL"))
  if (all(c("LCL", "UCL") %in% names(df))) return(c("LCL", "UCL"))
  NULL
}

ci_cols <- get_ci_cols(ct_df)

if (is.null(ci_cols)) {
  ct_df <- ct_df %>%
    mutate(
      lower = estimate - 1.96 * SE,
      upper = estimate + 1.96 * SE
    )
} else {
  ct_df <- ct_df %>%
    mutate(
      lower = .data[[ci_cols[1]]],
      upper = .data[[ci_cols[2]]]
    )
}

# --- determine significance: CI excludes 0 ---
plot_df <- ct_df %>%
  transmute(
    taskID,
    effect = estimate,          # independent − AI
    lower,
    upper,
    significant = (lower > 0 | upper < 0)
  ) %>%
  arrange(effect) %>%
  mutate(taskID = fct_inorder(taskID))

# --- forest plot ---
pdf("figure4a.pdf", width = 8, height = 6)
ggplot(plot_df, aes(x = effect, y = taskID)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.6) +
  geom_errorbarh(
    aes(xmin = lower, xmax = upper),
    height = 0.18,
    linewidth = 0.6
  ) +
  geom_point(
    aes(color = significant, shape = significant),
    size = 2.6
  ) +
  scale_color_manual(
    values = c("FALSE" = "grey50", "TRUE" = "black"),
    labels = c("Not significant", "Significant")
  ) +
  scale_shape_manual(
    values = c("FALSE" = 16, "TRUE" = 17),
    labels = c("Not significant", "Significant")
  ) +
  labs(
    x = "NASA-TLX difference (independent − AI)",
    y = NULL,
    color = NULL,
    shape = NULL,
    title = "Per-task NASA-TLX effect (independent − AI)"
  ) +
  theme_classic(base_size = 15) +
  theme(legend.position = "top")
dev.off()



```



## Figure 4b: forest plot for completion time difference, ordered by effect size (right)

```{r}
library(lme4)
library(emmeans)
library(dplyr)
library(ggplot2)
library(forcats)

completion_df <- completion_df %>%
  mutate(
    condition = factor(condition, levels = c("independent", "AI")),
    taskID    = factor(taskID)
  )

# --- model ---
m_task <- lmer(
  time ~ condition * taskID + (1 | participantID),
  data = completion_df
)

# --- emmeans contrast: independent - AI ---
emm <- emmeans(m_task, ~ condition | taskID)
ct  <- contrast(emm, method = list("independent - AI" = c(1, -1)), adjust = "none")

# --- summary with CIs (robust across emmeans versions) ---
ct_df <- as.data.frame(summary(ct, infer = c(TRUE, TRUE)))

get_ci_cols <- function(df) {
  if (all(c("lower.CL", "upper.CL") %in% names(df))) return(c("lower.CL", "upper.CL"))
  if (all(c("asymp.LCL", "asymp.UCL") %in% names(df))) return(c("asymp.LCL", "asymp.UCL"))
  if (all(c("LCL", "UCL") %in% names(df))) return(c("LCL", "UCL"))
  NULL
}

ci_cols <- get_ci_cols(ct_df)

if (is.null(ci_cols)) {
  ct_df <- ct_df %>%
    mutate(
      lower = estimate - 1.96 * SE,
      upper = estimate + 1.96 * SE
    )
} else {
  ct_df <- ct_df %>%
    mutate(
      lower = .data[[ci_cols[1]]],
      upper = .data[[ci_cols[2]]]
    )
}

# --- determine significance: CI excludes 0 ---
plot_df <- ct_df %>%
  transmute(
    taskID,
    effect = estimate,          # independent − AI
    lower,
    upper,
    significant = (lower > 0 | upper < 0)
  ) %>%
  arrange(effect) %>%
  mutate(taskID = fct_inorder(taskID))

# --- forest plot ---
pdf("figure4b.pdf", width = 8, height = 6)
ggplot(plot_df, aes(x = effect, y = taskID)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.6) +
  geom_errorbarh(
    aes(xmin = lower, xmax = upper),
    height = 0.18,
    linewidth = 0.6
  ) +
  geom_point(
    aes(color = significant, shape = significant),
    size = 2.6
  ) +
  scale_color_manual(
    values = c("FALSE" = "grey50", "TRUE" = "black"),
    labels = c("Not significant", "Significant")
  ) +
  scale_shape_manual(
    values = c("FALSE" = 16, "TRUE" = 17),
    labels = c("Not significant", "Significant")
  ) +
  labs(
    x = "completion time difference (independent − AI)",
    y = NULL,
    color = NULL,
    shape = NULL,
    title = "Completion Time Difference (independent − AI), in seconds "
  ) +
  theme_classic(base_size = 15) +
  theme(legend.position = "top")

dev.off()

```


### additional plots
# completion time across tasks
```{r}
library(dplyr)
library(ggplot2)
library(stringr)

# summarize mean time and standard error for each taskID
plot_df <- completion_df %>%
  mutate(
    CQ = str_extract(taskID, "^C\\dQ\\d"),
    G  = str_extract(taskID, "G\\d$")
  ) %>%
  group_by(CQ, G, taskID) %>%
  summarise(
    mean_time = mean(time, na.rm = TRUE),
    se_time   = sd(time, na.rm = TRUE) / sqrt(sum(!is.na(time))),
    .groups = "drop"
  ) %>%
  mutate(
    CQ = factor(
      CQ,
      levels = unique(str_sort(CQ, numeric = TRUE))
    ),
    G = factor(G, levels = c("G1", "G2"))
  )

ggplot(plot_df, aes(x = CQ, y = mean_time, fill = G)) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  geom_errorbar(
    aes(ymin = mean_time - se_time, ymax = mean_time + se_time),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  labs(
    x = "Task family (same C and Q)",
    y = "Average time",
    fill = "Difficulty",
    title = "Average time by task",
    subtitle = "G1 and G2 shown side by side within each C-Q task"
  ) +
  theme_minimal(base_size = 14)
```







```{r}
plot_df <- completion_df %>%
  mutate(
    CQ = str_extract(taskID, "^C\\dQ\\d"),
    G  = str_extract(taskID, "G\\d$")
  ) %>%
  group_by(CQ, G, taskID) %>%
  summarise(
    mean_time = mean(NASA_TLX_average, na.rm = TRUE),
    se_time   = sd(NASA_TLX_average, na.rm = TRUE) / sqrt(sum(!is.na(NASA_TLX_average))),
    .groups = "drop"
  ) %>%
  mutate(
    CQ = factor(
      CQ,
      levels = unique(str_sort(CQ, numeric = TRUE))
    ),
    G = factor(G, levels = c("G1", "G2"))
  )

ggplot(plot_df, aes(x = CQ, y = mean_time, fill = G)) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  geom_errorbar(
    aes(ymin = mean_time - se_time, ymax = mean_time + se_time),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  labs(
    x = "Task family (same C and Q)",
    y = "Average nasa-tlx",
    fill = "Difficulty",
    title = "Average nasa-tlx by task",
    subtitle = "G1 and G2 shown side by side within each C-Q task"
  ) +
  theme_minimal(base_size = 14)
```







